A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-16 latex/main.tex
Minor  edits

The LS HMM at the heart of sc2ts provides a simple and powerful means
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution
for the putative recombinant sequence
with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination
(similar to the idea of ``parsimony improvement'' in RIPPLES.)
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Properties of recombination events. A,B,C 
show summaries of the 354 sc2ts recombination nodes
passing quality control filters
colored by CovRecomb classification (recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table~\ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
Jackson group B (JB) and Jackson singleton recombinant 1 (J1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each QC-passing recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
for XA, XW, XBB, and an
undesignated recombination event under XBB, which we refer to as XBB.x.}
\label{fig:recombinant_evidence}
\end{figure}

@@

The 387 events detected by sc2ts and not by UShER+RIPPLES
would require a detailed dissection of the UShER tree
to explain, which is beyond the scope of this work.
However, the set does contain some well characterized events such as XBB.
We also compared the recombination events
output by RIPPLES with the Pango X lineages.
At p=4, we found that only XX of the Pango X lineages were present and
many well-characterized events such as XBB were not present.
We therefore reran at p=3, resulting in 4111 events overall,
and increasing the intersection of events with sc2ts to 595 (70\%).
At p=3, events involving XX of the 44 Pango X lineages
present in the dataset, with unequivocal origins for 10 lineages, and
7 lineages other lineages associated with two competing events.
[More summary]

>>>>

The LS HMM at the heart of sc2ts provides a simple and powerful means
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution
for the putative recombinant sequence
with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination.
[FIXME need to insert a sentence saying this is similar to 
metrics used by RIPPLES and other methods, contrasting
recombinant vs non-recombinant explanations.]
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Properties of recombination events. A,B,C 
show summaries of the 354 sc2ts recombination nodes
passing quality control filters
colored by 3SEQ classification (recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table~\ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
Jackson group B (JB) and Jackson singleton recombinant 1 (J1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x-y$).
(B) Distribution of time to the MRCA of the parents of each QC-passing recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
for XA, XW, XBB, and an
undesignated recombination event under XBB, which we refer to as XBB.x.}
\label{fig:recombinant_evidence}
\end{figure}

@@

The 387 events detected by sc2ts and not by UShER+RIPPLES
would require a detailed dissection of the UShER tree
to explain, which is beyond the scope of this work.
However, the set does contain some well characterized events such as XBB.
We also compared the recombination events
output by RIPPLES with the Pango X lineages.
At p=4, we found that only XX of the Pango X lineages were present and
many well-characterized events such as XBB were not present.
We therefore reran at p=3, resulting in 4111 events overall,
and increasing the intersection of events with sc2ts to 595 (70\%).
At p=3, events involving XX of the 44 Pango X lineages
are present in the dataset, with unequivocal origins for 10 lineages, and
7 lineages other lineages associated with two competing events.
[More summary]
